---
title: "Milestone2"
author: "pattiey"
output: github_document
---

# Pattie Ye's Mini Data Analysis: Milestone 2

First load the libraries

```{r message=FALSE}
library(datateachr)
library(tidyverse)
# I need the lubridate package to efficiently extract the date
library(lubridate)
# I need the ggridges package
library(ggridges)
```

# Task 1: Process and summarize data

## 1.1: My research questions

Research questions are changed from milestone 1 to be more specific. 

1. How has the frequency of trees planted by neighbourhood changed over time? And on a broader scale, how has it changed by decade? This could reflect the rate of development of certain neighbourhoods over time.

2. Is there a difference in the frequency of certain species of trees between neighbourhoods? What is the most popular species of tree within Vancouver and which neighbourhoods are these trees mostly found?

3. How does the number of trees planted change depending on the season?

4. Is there a relationship between the height and width of trees? Are these trees found in geographically similar areas or are they scattered throughout the city?

## 1.2 Summarizing and graphing

### Research Question 1

*How has the frequency of trees planted by neighbourhood changed over time? And on a broader scale, how has it changed by decade? This could reflect the rate of development of certain neighbourhoods over time.*

From Milestone 1, I created a variable called `year_planted` that contains the year in which a tree was planted.

```{r}
vancouver_trees <- vancouver_trees %>% 
  # use year() from lubridate to extract year from date
  mutate(year_planted = year(date_planted))
```

I just want the decade in which these trees were planted in order to get a better sense of overall trends over time. Create a categorical variable based on `year_planted`. 

```{r}
vancouver_trees <- vancouver_trees %>% 
  mutate(decade_planted = cut(year_planted, breaks = c(1990, 2000, 2010, 2020), labels = c("90's", "00's", "10's"))) 
```

I now want to see the count of trees planted per decade by neighbourhood.

```{r}
vancouver_trees %>% 
  group_by(neighbourhood_name, decade_planted) %>% 
  summarise(count = n())
```

Most of the trees do not contain information on `date_planted`. From the trees with date information, we see that most neighbourhoods saw the greatest number of trees planted between 2000-2010. This may suggest that the decade of 2000-2010 saw the greatest increase in development, but with large number of entries without date information, it is difficult to tell.

I would like to see the distribution of the rate of trees being planted (number of trees planted in a year) in a given neighbourhood and how that distribution changes depending on the decade. 

First, plot a histogram with `binwidth = 10` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 10) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 10", x = "Number of trees planted in a given year")
```

Now with `binwidth = 25` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 25) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 25", x = "Number of trees planted in a given year")
```

Finally, plot a histogram with `binwidth = 50` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 50) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 50", x = "Number of trees planted in a given year")
```

The histogram with `binwidth = 10` produces a graph where it is hard to see the differences between the decades and is difficult to extract anything meaningful from it. `binwidth = 50` does not give a great overview of the differences in distribution since the bins are quite wide. `binwidth = 25` provides a happy medium where you can get a sense of the distribution of the data without it being too difficult to read. Overlaying `geom_density()` gives an even better sense of the differences in distributions between the three decades. 

```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 25) +
  geom_density(alpha = 0.3) + 
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted in a neighbourhood", x = "Number of trees planted in a given year")
```

From this we see that the average rate of trees being planted per neighbourhood during the 2010's is lower than in the 1990's and 2000's. The 2000's saw the greatest average number of trees planted per year per neighbourhood. 

### Research Question 2

*Is there a difference in the frequency of certain species of trees between neighbourhoods? What is the most popular species of tree within Vancouver and which neighbourhoods are these trees mostly found?*

First, we shall look at summary statistics of the number of trees per type of tree in each neighbourhood. For each neighbourhood, find the mean, the median, min, and max number of trees per type, the most popular type of tree, and the total number of trees.

```{r}
vancouver_trees %>% 
  group_by(common_name, neighbourhood_name) %>% 
  # get a count of trees of each type per neighbourhood
  summarise(spec_count = n()) %>% 
  group_by(neighbourhood_name) %>% 
  arrange(desc(spec_count)) %>% 
  # names aren't as descriptive as I'd like in order to fit in the document
  summarise(mean = mean(spec_count), 
            median = median(spec_count), 
            min = min(spec_count), 
            max = max(spec_count),
            pop_tree = first(common_name), 
            total = sum(spec_count))
```

From this summary, we can get a general sense of which trees are most popular overall in the city (Kwanzan Flowering Cherry, Red Maple, Norway Maple). We can also tell that the distributions of numbers of trees per type is generally quite right-skewed (mean is greater than the median) which suggests that for most types of trees in each neighbourhood there are only a couple trees planted, while the popular trees are great in numbers. 

Now we look at the ten most popular types of trees in the city overall and see how the number of each of those trees per neighbourhood varies using a jitter plot overlaid with a violin plot.

```{r}
# first get the ten most popular trees in Vancouver
(most_pop_trees <- vancouver_trees %>% 
  group_by(common_name) %>% 
  summarise(spec_count = n()) %>% 
  arrange(desc(spec_count)) %>% 
  pull(common_name) %>% 
  head(10))

vancouver_trees %>% 
  # filter to keep only the most popular trees
  filter(common_name %in% most_pop_trees) %>% 
  group_by(common_name, neighbourhood_name) %>% 
  summarise(spec_count = n()) %>% 
  # Have tree names as a factor so that the graph is in order of popularity
  ggplot(aes(factor(common_name, levels = most_pop_trees), spec_count, color = common_name, fill = common_name)) +
  # jitter plot, we want to plot the actual number of trees, so eliminate vertical jitter
  geom_jitter(height = 0) +
  geom_violin(alpha = 0.2) +
  # Do not include legend since it is redundant
  theme(legend.position = "none") +
  labs(title = "Per neighbourhood count of most \npopular trees across Vancouver", 
       x = "Common name of tree", 
       y = "Count by neighbourhood") +
  coord_flip()
```

From this graph, we see that some types of trees such as the Kwanzan Flowering Cherry are quite popular across a number of neighbourhoods, while others such as the Red Maple are generally not as popular, but are prominent in a few neighbourhoods. 

### Research Question 3

*How does the number of trees planted change depending on the season?*

We need to create a categorical variable `season_planted`. The value of the variable will be assigned as so:

| `season_planted` | Months | 
| --- | --- |
| `winter` | Dec. (12), Jan. (1), Feb. (2)  | 
| `spring` | Mar. (3), Apr. (4), May (5) | 
| `summer` | Jun. (6), Jul. (7), Aug. (8) | 
| `autumn` | Sep. (9), Oct. (10), Nov. (11) |

```{r}
vancouver_trees <- vancouver_trees %>% 
  mutate(season_planted = case_when(
    month(date_planted) %in% 3:5 ~ "Spring",
    month(date_planted) %in% 6:8 ~ "Summer",
    month(date_planted) %in% 9:11 ~ "Autumn",
    is.na(month(date_planted)) ~ "NA",
    # if the month is not in any of those other ranges, it must be Winter
    TRUE ~ "Winter"
  ))

vancouver_trees %>% 
  group_by(season_planted) %>% 
  summarise(count = n())
```

Most of the trees do not have date information. From the trees that do have `date_planted` information, it is shown that Winter is the most popular season for trees to be planted, followed by Spring, Autumn, and lastly (by a lot) Summer.

Here, we plot the distribution of trees planted by date coloured by their season. 

Start with `binwidth = 1`, each bin is one day.

```{r}
vancouver_trees %>% 
  # Filter out trees without date information
  filter(!is.na(date_planted)) %>% 
  # Extract just the day of the year from date_planted
  mutate(day = yday(date_planted)) %>% 
  ggplot(aes(day, fill = season_planted)) + 
  geom_histogram(binwidth = 1) + 
  labs(x = "Day of the year", 
       fill = "Season planted", 
       title = "Distribution of trees planted \nby day of the year, binwidth = 1")
```
From this, we see that there is one day in Autumn that appears to be an outlier with close to 800 trees planted in a single day. We also see that towards the Summer, as the weather starts getting warmer, tree planting slows down.

Now plot with `binwidth = 7`. Each bin is a week.

```{r}
vancouver_trees %>% 
  # Filter out trees without date information
  filter(!is.na(date_planted)) %>% 
  # Extract just the day of the year from date_planted
  mutate(day = yday(date_planted)) %>% 
  ggplot(aes(day, fill = season_planted)) + 
  geom_histogram(binwidth = 7) + 
  labs(x = "Day of the year", 
       fill = "Season planted", 
       title = "Distribution of trees planted \nby day of the year, binwidth = 7")
```

From this histogram we see that the busiest weeks of the year for tree planting are between Winter and Spring, and late Autumn. It is also interesting to see that during the week surrounding the holiday season in late December/early January, we see a dramatic decrease in the number of trees planted. This histogram is a bit smoother and easier to read. 

Now plot with `bins = 12`. Each bin roughly represents one month. 

```{r}
vancouver_trees %>% 
  # Filter out trees without date information
  filter(!is.na(date_planted)) %>% 
  # Extract just the day of the year from date_planted
  mutate(day = yday(date_planted)) %>% 
  ggplot(aes(day, fill = season_planted)) + 
  geom_histogram(bins = 12) + 
  labs(x = "Day of the year", 
       fill = "Season planted", 
       title = "Distribution of trees planted \nby day of the year, bins = 12")
```

From this histogram, we see that February, March/April, and November/December are the busiest months for tree planting. Although, this graph does not cleanly divide up the months, if we are looking for the trees by month, it would be better to directly plot a histogram using the month. 

```{r}
vancouver_trees %>% 
  # Filter out trees without date information
  filter(!is.na(date_planted)) %>% 
  # Extract just the day of the year from date_planted
  mutate(month = month(date_planted)) %>% 
  ggplot(aes(month, fill = season_planted)) + 
  geom_histogram(binwidth = 1) + 
  labs(x = "Month planted", 
       fill = "Season planted", 
       title = "Distribution of trees planted \nby month")
```

Here, see that the busiest tree planting months are actually January, February, March, and November, which is difficult to tell from the previous histogram. 

The most informative histogram in this case would be that one with `binwidth = 7` since it is smoother than the first histogram, eliminates the effect of the outlying day, while also being able to show interesting trends in planting such as the decrease surrounding the holiday season and the steady decrease and increase in planting surrounding the Summer.

### Research Question 4

*Is there a relationship between the height and width of trees? Are these trees found in geographically similar areas or are they scattered throughout the city?*

Since `height_range_id` is an ID variable, it can be viewed as a categorical variable. Let's look at the summary statistics of `diameter` across the different `height_range_id` values. 

For each `height_range_id` find the range, mean, median, first and third quartile values, as well as what type of tree the thinnest and thickest tree are.

```{r}
vancouver_trees %>% 
  group_by(height_range_id) %>% 
  summarise(min = min(diameter), 
            first_q = quantile(diameter, probs = 0.25), 
            median = median(diameter), 
            mean = mean(diameter), 
            third_q = quantile(diameter, probs = 0.75), 
            max = max(diameter))
# For visual purposes, thinnest and thickest types are separate
vancouver_trees %>% 
  group_by(height_range_id) %>%
  arrange(diameter) %>% 
  summarise(thinnest_type = first(common_name), thickest_type = last(common_name))
```

Based on the median and mean values, it appears that a higher `height_range_id` corresponds to thicker trees. But it is interesting to note that the maximum tree diameters belong in `height_range_id` 3, 4, and 5. There doesn't seem to be a clear explanation for these outliers since they are much greater than the other diameter values. Perhaps there was a recording error.

Here, we plot the diameter against `height_range_id` using a jitter plot.

```{r}
vancouver_trees %>% ggplot(aes(factor(height_range_id, levels = 1:10), diameter)) + geom_jitter(height = 0, alpha = 0.1) + labs(y = "Diameter", x = "Height range ID", title = "Tree diameter vs. Height range ID")
```

This is extremely difficult to read. Use a $$log_{10}$$ transform on the y-axis. 

```{r}
vancouver_trees %>% ggplot(aes(factor(height_range_id, levels = 1:10), diameter)) + geom_jitter(height = 0, alpha = 0.2) + scale_y_continuous(trans = 'log10') + labs(y = "Diameter", x = "Height range ID", title = "Tree diameter vs. Height range ID")
```
This still is not great, but is better. You can see that as the `height_range_id` increases, the number of smaller diameter trees decreases. Perhaps using a box plot will give something more interpretable.

```{r}
vancouver_trees %>% 
  ggplot(aes(factor(height_range_id, levels = 1:10), diameter)) + 
  geom_boxplot(alpha = 0.2) + 
  scale_y_continuous(trans = 'log10') + 
  labs(y = "Diameter", 
       x = "Height range ID", 
       title = "Box plot of tree diameter vs. height range ID")
```

Indeed it does. From this, it is clear that as the height increases, the average tree diameter also increases.

## 1.3 Assess research questions


