---
title: "Milestone1"
author: "pattiey"
output: github_document
---
# Pattie Ye's Mini Data Analysis: Milestone 1

# Choosing a dataset from the `datateachr` package

First load the libraries

```{r}
suppressMessages(library(datateachr))
suppressMessages(library(tidyverse))
# I need the lubridate package to efficiently extract the date
suppressMessages(library(lubridate))
# I need the ggridges package
suppressMessages(library(ggridges))
```

From the 7 datasets in `datateachr` my top 4 choices are:

1. `cancer_sample`: characteristics of cell nuclei from digitized images of breast mass

```{r}
# Look at column names
names(cancer_sample)

# Dimensions
dim(cancer_sample)

# Data class of each column
sapply(cancer_sample, class)
```

2. `vancouver_trees`: listing of public trees on boulevards in the City of Vancouver 

```{r}
# Look at column names
names(vancouver_trees)

# Dimensions
dim(vancouver_trees)

# Data type of each column
sapply(vancouver_trees, class)
```

3. `flow_sample`: Maximum and minimum recorded mean daily flow recorded at the Bow River Station in Banff 

```{r}
# Look at column names
names(flow_sample)

# Dimensions
dim(flow_sample)

# Data type of each column
sapply(flow_sample, class)
```

4. `parking_meters`: information on rates and time limits for parking meters in the City of Vancouver 

```{r}
# Look at column names
names(parking_meters)

# Dimensions
dim(parking_meters)

# Data type of each column
sapply(parking_meters, class)
```

# Top 2 choices:

* `vancouver_trees`
* `flow_sample`

## `vancouver_trees`

```{r}
# Looking at vancouver_trees to get an idea of values
glimpse(vancouver_trees)

# Look at how many unique values exist in each feature in vancouver_trees
vancouver_trees %>% sapply(unique) %>% sapply(length)

# The height_range_id variable is ambiguous, look at what values it takes on 
unique(vancouver_trees$height_range_id)
# height range is between 0 and 10

# Look at all fields with only 2 unique values
unique(vancouver_trees$assigned)
unique(vancouver_trees$root_barrier)
unique(vancouver_trees$curb)
# They all seem to be Y/N (boolean) type features

# What is street side name describing?
unique(vancouver_trees$street_side_name)
# Where on the street is the tree located on

# I'm not too sure what plant_area is. 
unique(vancouver_trees$plant_area)
```

`vancouver_trees` describes information about public trees planted on boulevards in the City of Vancouver. This dataset interests me because there are so many different types of information available, such as location, type of tree, and date planted. 
With this data, I could answer a variety of different questions, such as:

* What is the location of trees planted in Vancouver?
* How has the number of trees planted varied from year to year?
* How do trees vary physically depending on tree type? 

## `flow_sample`

```{r}
# Looking at flow_sample to get an idea of values 
glimpse(flow_sample)

# How many unique values exist for each feature
flow_sample %>% sapply(unique) %>% sapply(length)

# Confirm that extreme_type is either max or min
unique(flow_sample$extreme_type)

# How is it that there are 32 unique values for day
unique(flow_sample$day)
# there are NA values... how many?
sum(is.na(flow_sample$day))
```

`flow_sample` contains information about the daily flow data for the Bow River at Banff. Although the data is relatively simple, it can answer many interesting questions. Some questions that can be answered are:

* How flow has changed over time?
* During what months do the extremes typically happen? 

# Final choice

`vancouver_trees` seems more interesting to play around with since there are so many different types of features that can demonstrate a large variety of information. 

# Exploring My Dataset

I'm interested in seeing how the number of trees planted varies from year to year, so I need a variable that contains only the year from `date_planted` called `year_planted`

```{r}
vancouver_trees <- vancouver_trees %>% 
  # use year() from lubridate to extract year from date
  mutate(year_planted = year(date_planted))
```

I would like to see how the number of trees planted changed year to year

```{r}
ggplot(vancouver_trees, aes(year_planted)) + geom_density() + theme_classic() + xlab("Year") + ggtitle("Density of trees planted per year")
```

I now want to see how that distribution changes by neighbourhood

```{r}
ggplot(vancouver_trees, aes(x = year_planted, y = neighbourhood_name)) + geom_density_ridges() + xlab("Year") + ylab("Neighbourhood") + ggtitle("Density of trees planted per year, by neighbourhood")
```

I'd also like to see how many trees there are in each neighbourhood.

```{r}
ggplot(vancouver_trees, aes(neighbourhood_name)) + geom_bar() + theme(axis.text.x = element_text(angle = 90)) + xlab("Neighbourhood") + ggtitle("Number of trees planted per neighbourhood")
```

This is sort of hard to interpret as a graph, I'd like to just see it as a table instead.

```{r}
vancouver_trees %>% group_by(neighbourhood_name) %>% count()
```

I'm most interested in the trees in my neighbourhood, Kitsilano, where there are 8115 trees. I only care about trees that are greater than 25 and less than 100 in diameter. I want information about the location first and then the rest since I would like to know the geographic location of these trees.

```{r}
kits_trees <- vancouver_trees %>% 
  filter(neighbourhood_name == "KITSILANO") %>% 
  select(on_street, on_street_block, street_side_name, curb, longitude, latitude, everything())
```

I want to see where in kits trees between 25 and 100 in diameter are planted and where on the street they are.

```{r}
kits_trees %>% 
  filter(diameter >= 25, diameter <= 100) %>% 
  ggplot(aes(longitude, latitude)) + 
  geom_point(aes(col = street_side_name, size = diameter), alpha = 0.2) + 
  ggtitle("Location of trees in Kitsilano") + 
  labs(color = "Street Side") + labs(size = "Tree Diameter")
```

There appears to be one outlying tree in longitude, I want to filter that out.

```{r}
kits_trees %>% 
  filter(diameter >= 25, diameter <= 100, longitude < -123.1) %>% 
  ggplot(aes(longitude, latitude)) + 
  geom_point(aes(col = street_side_name, size = diameter), alpha = 0.2) + 
  ggtitle("Location of trees in Kitsilano") + 
  labs(color = "Street Side") + labs(size = "Tree Diameter")
```

# Research questions

* How have tree planting patterns changed over years?

* What are the most popular species of trees in Vancouver, does that change by neighbourhood?

* Which streets in Vancouver are most densely populated with trees?

* Where can you find the largest (tallest and widest) trees in Vancouver?

