---
title: "Milestone2"
author: "pattiey"
output: github_document
---

# Pattie Ye's Mini Data Analysis: Milestone 2

First load the libraries

```{r message=FALSE}
library(datateachr)
library(tidyverse)
# I need the lubridate package to efficiently extract the date
library(lubridate)
# I need the ggridges package
library(ggridges)
```

# Task 1: Process and summarize data

## 1.1: My research questions

Research questions are changed from milestone 1 to be more specific. 

1. How has the frequency of trees planted by neighbourhood changed over time? And on a broader scale, how has it changed by decade? This could reflect the rate of development of certain neighbourhoods over time.

2. Is there a difference in the frequency of certain species of trees between neighbourhoods? What is the most popular species of tree within Vancouver and which neighbourhoods are these trees mostly found?

3. How does the number of trees planted change depending on the season? Do we see a difference in planting season between popular species of trees?

4. Is there a relationship between the height and width of trees? Are these trees found in geographically similar areas or are they scattered throughout the city?

## 1.2 Summarizing and graphing

### Research Question 1

*How has the frequency of trees planted by neighbourhood changed over time? And on a broader scale, how has it changed by decade? This could reflect the rate of development of certain neighbourhoods over time.*

From Milestone 1, I created a variable called `year_planted` that contains the year in which a tree was planted.

```{r}
vancouver_trees <- vancouver_trees %>% 
  # use year() from lubridate to extract year from date
  mutate(year_planted = year(date_planted))
```

I just want the decade in which these trees were planted in order to get a better sense of overall trends over time. Create a categorical variable based on `year_planted`. 

```{r}
vancouver_trees <- vancouver_trees %>% 
  mutate(decade_planted = cut(year_planted, breaks = c(1990, 2000, 2010, 2020), labels = c("90's", "00's", "10's"))) 
```

I now want to see the count of trees planted per decade by neighbourhood.

```{r}
vancouver_trees %>% 
  group_by(neighbourhood_name, decade_planted) %>% 
  summarise(count = n())
```

Most of the trees do not contain information on `date_planted`. From the trees with date information, we see that most neighbourhoods saw the greatest number of trees planted between 2000-2010. This may suggest that the decade of 2000-2010 saw the greatest increase in development, but with large number of entries without date information, it is difficult to tell.

I would like to see the distribution of the rate of trees being planted (number of trees planted in a year) in a given neighbourhood and how that distribution changes depending on the decade. 

First, plot a histogram with `binwidth = 10` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 10) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 10", x = "Number of trees planted in a given year")
```

Now with `binwidth = 25` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 25) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 25", x = "Number of trees planted in a given year")
```

Finally, plot a histogram with `binwidth = 50` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 50) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 50", x = "Number of trees planted in a given year")
```

The histogram with `binwidth = 10` produces a graph where it is hard to see the differences between the decades and is difficult to extract anything meaningful from it. `binwidth = 50` does not give a great overview of the differences in distribution since the bins are quite wide. `binwidth = 25` provides a happy medium where you can get a sense of the distribution of the data without it being too difficult to read. Overlaying `geom_density()` gives an even better sense of the differences in distributions between the three decades. 

```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 25) +
  geom_density(alpha = 0.3) + 
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted in a neighbourhood", x = "Number of trees planted in a given year")
```

From this we see that the average rate of trees being planted per neighbourhood during the 2010's is lower than in the 1990's and 2000's. The 2000's saw the greatest average number of trees planted per year per neighbourhood. 

### Research Question 2

*Is there a difference in the frequency of certain species of trees between neighbourhoods? What is the most popular species of tree within Vancouver and which neighbourhoods are these trees mostly found?*

First, we shall look at summary statistics of the number of trees per type of tree in each neighbourhood. For each neighbourhood, find the mean, the median, min, and max number of trees per type, the most popular type of tree, and the total number of trees.

```{r}
vancouver_trees %>% 
  group_by(common_name, neighbourhood_name) %>% 
  # get a count of trees of each type per neighbourhood
  summarise(spec_count = n()) %>% 
  group_by(neighbourhood_name) %>% 
  arrange(desc(spec_count)) %>% 
  # names aren't as descriptive as I'd like in order to fit in the document
  summarise(mean = mean(spec_count), 
            median = median(spec_count), 
            min = min(spec_count), 
            max = max(spec_count),
            pop_tree = first(common_name), 
            total = sum(spec_count))
```

From this summary, we can get a general sense of which trees are most popular overall in the city (Kwanzan Flowering Cherry, Red Maple, Norway Maple). We can also tell that the distributions of numbers of trees per type is generally quite right-skewed (mean is greater than the median) which suggests that for most types of trees in each neighbourhood there are only a couple trees planted, while the popular trees are great in numbers. 

Now we look at the ten most popular types of trees in the city overall and see how the number of each of those trees per neighbourhood varies using a jitter plot overlaid with a violin plot.

```{r}
# first get the ten most popular trees in Vancouver
(most_pop_trees <- vancouver_trees %>% 
  group_by(common_name) %>% 
  summarise(spec_count = n()) %>% 
  arrange(desc(spec_count)) %>% 
  pull(common_name) %>% 
  head(10))

vancouver_trees %>% 
  # filter to keep only the most popular trees
  filter(common_name %in% most_pop_trees) %>% 
  group_by(common_name, neighbourhood_name) %>% 
  summarise(spec_count = n()) %>% 
  ggplot(aes(factor(common_name, levels = most_pop_trees), spec_count, color = common_name, fill = common_name)) +
  # jitter plot, we want to plot the actual number of trees, so eliminate vertical jitter
  geom_jitter(height = 0) +
  geom_violin(alpha = 0.2) +
  # Do not include legend since it is redundant
  theme(legend.position = "none") +
  labs(title = "Per neighbourhood count of most \n popular trees across Vancouver", 
       x = "Common name of tree", 
       y = "Count by neighbourhood") +
  coord_flip()
```

From this graph, we see that some types of trees such as the Kwanzan Flowering Cherry are quite popular across a number of neighbourhoods, while 
