---
title: "Milestone2"
author: "pattiey"
output: github_document
---

# Pattie Ye's Mini Data Analysis: Milestone 2

First load the libraries

```{r message=FALSE}
library(datateachr)
library(tidyverse)
# I need the lubridate package to efficiently extract the date
library(lubridate)
# I need the ggridges package
library(ggridges)
```

# Task 1: Process and summarize data

## 1.1: My research questions

Research questions are changed from milestone 1 to be more specific. 

1. How has the frequency of trees planted by neighbourhood changed over time? And on a broader scale, how has it changed by decade? This could reflect the rate of development of certain neighbourhoods over time.

2. Is there a difference in the frequency of certain species of trees between neighbourhoods? What are the most popular species of tree within Vancouver and which neighbourhoods are these trees mostly found?

3. How does the number of trees planted change depending on the season?

4. Is there a relationship between the height and width of trees? Are these larger trees found in geographically similar areas or are they scattered throughout the city?

## 1.2 Summarizing and graphing

### Research Question 1

*How has the frequency of trees planted by neighbourhood changed over time? And on a broader scale, how has it changed by decade? This could reflect the rate of development of certain neighbourhoods over time.*

From Milestone 1, I created a variable called `year_planted` that contains the year in which a tree was planted.

```{r}
vancouver_trees <- vancouver_trees %>% 
  # use year() from lubridate to extract year from date
  mutate(year_planted = year(date_planted))
```

I just want the decade in which these trees were planted in order to get a better sense of overall trends over time. Create a categorical variable based on `year_planted`. 

```{r}
vancouver_trees <- vancouver_trees %>% 
  mutate(decade_planted = cut(year_planted, breaks = c(1990, 2000, 2010, 2020), labels = c("90's", "00's", "10's"))) 
```

I now want to see the count of trees planted per decade by neighbourhood.

```{r}
vancouver_trees %>% 
  group_by(neighbourhood_name, decade_planted) %>% 
  summarise(count = n())
```

Most of the trees do not contain information on `date_planted`. From the trees with date information, we see that most neighbourhoods saw the greatest number of trees planted between 2000-2010. This may suggest that the decade of 2000-2010 saw the greatest increase in development, but with large number of entries without date information, it is difficult to tell.

I would like to see the distribution of the rate of trees being planted (number of trees planted in a year) in a given neighbourhood and how that distribution changes depending on the decade. 

First, plot a histogram with `binwidth = 10` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 10) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 10", x = "Number of trees planted in a given year")
```

Now with `binwidth = 25` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 25) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 25", x = "Number of trees planted in a given year")
```

Finally, plot a histogram with `binwidth = 50` 
```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 50) +
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted, bin = 50", x = "Number of trees planted in a given year")
```

The histogram with `binwidth = 10` produces a graph where it is hard to see the differences between the decades and is difficult to extract anything meaningful from it. `binwidth = 50` does not give a great overview of the differences in distribution since the bins are quite wide. `binwidth = 25` provides a happy medium where you can get a sense of the distribution of the data without it being too difficult to read. Overlaying `geom_density()` gives an even better sense of the differences in distributions between the three decades. 

```{r}
vancouver_trees %>% 
  group_by(year_planted, neighbourhood_name, decade_planted) %>% 
  summarise(annual_trees_planted = n()) %>% 
  # omit trees that do not have date planted information 
  filter(!is.na(decade_planted)) %>%  
  ggplot(aes(annual_trees_planted, fill = decade_planted, color = decade_planted)) +
  geom_histogram(position = "dodge", alpha = 0.2, aes(y = ..density..), binwidth = 25) +
  geom_density(alpha = 0.3) + 
  labs(fill = "Decade Planted", color = "Decade Planted", title = "Density distribution of annual number of trees planted in a neighbourhood", x = "Number of trees planted in a given year")
```

From this we see that the average rate of trees being planted per neighbourhood during the 2010's is lower than in the 1990's and 2000's. The 2000's saw the greatest average number of trees planted per year per neighbourhood. 

### Research Question 2

*Is there a difference in the frequency of certain species of trees between neighbourhoods? What are the most popular species of tree within Vancouver and which neighbourhoods are these trees mostly found?*

First, we shall look at summary statistics of the number of trees per type of tree in each neighbourhood. For each neighbourhood, find the mean, the median, min, and max number of trees per type, the most popular type of tree, and the total number of trees.

```{r}
vancouver_trees %>% 
  group_by(common_name, neighbourhood_name) %>% 
  # get a count of trees of each type per neighbourhood
  summarise(spec_count = n()) %>% 
  group_by(neighbourhood_name) %>% 
  arrange(desc(spec_count)) %>% 
  # names aren't as descriptive as I'd like in order to fit in the document
  summarise(mean = mean(spec_count), 
            median = median(spec_count), 
            min = min(spec_count), 
            max = max(spec_count),
            pop_tree = first(common_name), 
            total = sum(spec_count))
```

From this summary, we can get a general sense of which trees are most popular overall in the city (Kwanzan Flowering Cherry, Red Maple, Norway Maple). We can also tell that the distributions of numbers of trees per type is generally quite right-skewed (mean is greater than the median) which suggests that for most types of trees in each neighbourhood there are only a couple trees planted, while the popular trees are great in numbers. 

Now we look at the ten most popular types of trees in the city overall and see how the number of each of those trees per neighbourhood varies using a jitter plot overlaid with a violin plot.

```{r}
# first get the ten most popular trees in Vancouver
(most_pop_trees <- vancouver_trees %>% 
  group_by(common_name) %>% 
  summarise(spec_count = n()) %>% 
  arrange(desc(spec_count)) %>% 
  pull(common_name) %>% 
  head(10))

vancouver_trees %>% 
  # filter to keep only the most popular trees
  filter(common_name %in% most_pop_trees) %>% 
  group_by(common_name, neighbourhood_name) %>% 
  summarise(spec_count = n()) %>% 
  # Have tree names as a factor so that the graph is in order of popularity
  ggplot(aes(factor(common_name, levels = most_pop_trees), spec_count, color = common_name, fill = common_name)) +
  # jitter plot, we want to plot the actual number of trees, so eliminate vertical jitter
  geom_jitter(height = 0) +
  geom_violin(alpha = 0.2) +
  # Do not include legend since it is redundant
  theme(legend.position = "none") +
  labs(title = "Per neighbourhood count of most \npopular trees across Vancouver", 
       x = "Common name of tree", 
       y = "Count by neighbourhood") +
  coord_flip()
```

From this graph, we see that some types of trees such as the Kwanzan Flowering Cherry are quite popular across a number of neighbourhoods, while others such as the Red Maple are generally not as popular, but are prominent in a few neighbourhoods. 

### Research Question 3

*How does the number of trees planted change depending on the season?*

We need to create a categorical variable `season_planted`. The value of the variable will be assigned as so:

| `season_planted` | Months | 
| --- | --- |
| `winter` | Dec. (12), Jan. (1), Feb. (2)  | 
| `spring` | Mar. (3), Apr. (4), May (5) | 
| `summer` | Jun. (6), Jul. (7), Aug. (8) | 
| `autumn` | Sep. (9), Oct. (10), Nov. (11) |

```{r}
vancouver_trees <- vancouver_trees %>% 
  mutate(season_planted = case_when(
    month(date_planted) %in% 3:5 ~ "Spring",
    month(date_planted) %in% 6:8 ~ "Summer",
    month(date_planted) %in% 9:11 ~ "Autumn",
    is.na(month(date_planted)) ~ "NA",
    # if the month is not in any of those other ranges, it must be Winter
    TRUE ~ "Winter"
  ))

vancouver_trees %>% 
  group_by(season_planted) %>% 
  summarise(count = n())
```

Most of the trees do not have date information. From the trees that do have `date_planted` information, it is shown that Winter is the most popular season for trees to be planted, followed by Spring, Autumn, and lastly (by a lot) Summer.

Here, we plot the distribution of trees planted by date coloured by their season. 

Start with `binwidth = 1`, each bin is one day.

```{r}
vancouver_trees %>% 
  # Filter out trees without date information
  filter(!is.na(date_planted)) %>% 
  # Extract just the day of the year from date_planted
  mutate(day = yday(date_planted)) %>% 
  ggplot(aes(day, fill = season_planted)) + 
  geom_histogram(binwidth = 1) + 
  labs(x = "Day of the year", 
       fill = "Season planted", 
       title = "Distribution of trees planted \nby day of the year, binwidth = 1")
```

From this, we see that there is one day in Autumn that appears to be an outlier with close to 800 trees planted in a single day. We also see that towards the Summer, as the weather starts getting warmer, tree planting slows down.

Now plot with `binwidth = 7`. Each bin is a week.

```{r}
vancouver_trees %>% 
  # Filter out trees without date information
  filter(!is.na(date_planted)) %>% 
  # Extract just the day of the year from date_planted
  mutate(day = yday(date_planted)) %>% 
  ggplot(aes(day, fill = season_planted)) + 
  geom_histogram(binwidth = 7) + 
  labs(x = "Day of the year", 
       fill = "Season planted", 
       title = "Distribution of trees planted \nby day of the year, binwidth = 7")
```

From this histogram we see that the busiest weeks of the year for tree planting are between Winter and Spring, and late Autumn. It is also interesting to see that during the week surrounding the holiday season in late December/early January, we see a dramatic decrease in the number of trees planted. This histogram is a bit smoother and easier to read. 

Now plot with `bins = 12`. Each bin roughly represents one month. 

```{r}
vancouver_trees %>% 
  # Filter out trees without date information
  filter(!is.na(date_planted)) %>% 
  # Extract just the day of the year from date_planted
  mutate(day = yday(date_planted)) %>% 
  ggplot(aes(day, fill = season_planted)) + 
  geom_histogram(bins = 12) + 
  labs(x = "Day of the year", 
       fill = "Season planted", 
       title = "Distribution of trees planted \nby day of the year, bins = 12")
```

From this histogram, we see that February, March/April, and November/December are the busiest months for tree planting. Although, this graph does not cleanly divide up the months, if we are looking for the trees by month, it would be better to directly plot a histogram using the month. 

```{r}
vancouver_trees %>% 
  # Filter out trees without date information
  filter(!is.na(date_planted)) %>% 
  # Extract just the day of the year from date_planted
  mutate(month = month(date_planted)) %>% 
  ggplot(aes(month, fill = season_planted)) + 
  geom_histogram(binwidth = 1) + 
  labs(x = "Month planted", 
       fill = "Season planted", 
       title = "Distribution of trees planted \nby month")
```

Here, see that the busiest tree planting months are actually January, February, March, and November, which is difficult to tell from the previous histogram. 

The most informative histogram in this case would be that one with `binwidth = 7` since it is smoother than the first histogram, eliminates the effect of the outlying day, while also being able to show interesting trends in planting such as the decrease surrounding the holiday season and the steady decrease and increase in planting surrounding the Summer.

### Research Question 4

*Is there a relationship between the height and width of trees? Are these larger trees found in geographically similar areas or are they scattered throughout the city?*

Since `height_range_id` is an ID variable, it can be viewed as a categorical variable. Let's look at the summary statistics of `diameter` across the different `height_range_id` values. 

For each `height_range_id` find the range, mean, median, first and third quartile values, as well as what type of tree the thinnest and thickest tree are.

```{r}
vancouver_trees %>% 
  group_by(height_range_id) %>% 
  summarise(min = min(diameter), 
            first_q = quantile(diameter, probs = 0.25), 
            median = median(diameter), 
            mean = mean(diameter), 
            third_q = quantile(diameter, probs = 0.75), 
            max = max(diameter))
# For visual purposes, thinnest and thickest types are separate
vancouver_trees %>% 
  group_by(height_range_id) %>%
  arrange(diameter) %>% 
  summarise(thinnest_type = first(common_name), thickest_type = last(common_name))
```

Based on the median and mean values, it appears that a higher `height_range_id` corresponds to thicker trees. But it is interesting to note that the maximum tree diameters belong in `height_range_id` 3, 4, and 5. There doesn't seem to be a clear explanation for these outliers since they are much greater than the other diameter values. Perhaps there was a recording error.

Here, we plot the diameter against `height_range_id` using a jitter plot.

```{r}
vancouver_trees %>% ggplot(aes(factor(height_range_id, levels = 0:10), diameter)) + geom_jitter(height = 0, alpha = 0.1) + labs(y = "Diameter", x = "Height range ID", title = "Tree diameter vs. Height range ID")
```

This is extremely difficult to read. Use a $$log_{10}$$ transform on the y-axis. 

```{r}
vancouver_trees %>% ggplot(aes(factor(height_range_id, levels = 0:10), diameter)) + geom_jitter(height = 0, alpha = 0.2) + scale_y_continuous(trans = 'log10') + labs(y = "Diameter", x = "Height range ID", title = "Tree diameter vs. Height range ID")
```
This still is not great, but is better. You can see that, generally, as the `height_range_id` increases, the number of smaller diameter trees decreases. Perhaps using a box plot will give something more interpretable.

```{r}
vancouver_trees %>% 
  ggplot(aes(factor(height_range_id, levels = 0:10), diameter)) + 
  geom_boxplot(alpha = 0.2) + 
  scale_y_continuous(trans = 'log10') + 
  labs(y = "Diameter", 
       x = "Height range ID", 
       title = "Box plot of tree diameter vs. height range ID")
```

Indeed it does. From this, it is clear that in general, as the height increases, the average tree diameter also increases.

For fun, I want to see if the presence of a root barrier has any relationship with diameter. 

```{r}
vancouver_trees %>% 
  ggplot(aes(factor(height_range_id, levels = 0:10), diameter, color = root_barrier)) + 
  geom_boxplot(alpha = 0.2) + 
  scale_y_continuous(trans = 'log10') + 
  labs(y = "Diameter", 
       x = "Height range ID", 
       title = "Box plot of tree diameter vs. height range ID",
       color = "Root barrier")
```

In general, it appears that the presence of a root barrier corresponds to a lower tree diameter.

## 1.3 Assess research questions

### Research Question 1

*How has the frequency of trees planted by neighbourhood changed over time? And on a broader scale, how has it changed by decade? This could reflect the rate of development of certain neighbourhoods over time.*

I am closer to answering this question in that I have identified the different distributions of number of trees planted per year in each neighbourhood. From this we see that the 2000's had the highest average number of neighbourhood trees planted annually. What remains unclear is which neighbourhoods saw the greatest amounts of growth in each decade and which neighbourhoods had the biggest variation of trees planted between decades. I'd like to refine my research question to be: which neighbourhoods saw the greatest rate of trees planted per decade? Are there any neighbourhoods that have only recently seen a high rate of tree planting (last 5 years?) and conversely, are there any neighbourhoods that have drastically slowed down in their rate of tree planting?

### Research Question 2

*Is there a difference in the frequency of certain species of trees between neighbourhoods? What are the most popular species of tree within Vancouver and which neighbourhoods are these trees mostly found?*

I have identified the most popular types of trees within each neighbourhood and throughout Vancouver as a whole. What remains to be answered is the specific neighbourhoods where the most popular trees are mainly found. It would also be interesting to see if all the popular trees are concentrated in one area in the neighbourhood or are scattered around.

### Research Question 3

*How does the number of trees planted change depending on the season?*

I think this question has been sufficiently answered. The number of trees planted decreases as the weather gets warmer in the Summer months, as well as in late December/early January surrounding the holiday season. It may be interesting to create a time series model for the number of trees planted through time.

### Research Question 4

*Is there a relationship between the height and width of trees? Are these larger trees found in geographically similar areas or are they scattered throughout the city?*

From the data exploration, we found that on average, a larger `height_range_id` corresponds to a larger diameter, which answers the first part of the question. The second part of the research remains unanswered. It would be interesting to explore the geographic relationship between the trees that are taller (ex. `height_range_id` > 5) and wider in diameter (ex. `diameter > 20`). It would also be interesting to see if the age of a tree (based on `date_planted`) has any relationship with the height or diameter of a tree.

# Task 2: Tidy my data

## 2.1 Is my data tidy?

I believe my data is tidy. Each row represents one tree (one observation), each column represents a different variable describing the tree, and each cell represents the value corresponding to that tree. To be sure, let's go through the 8 variables:

* `tree_id`
* `std_street`
* `common_name`
* `height_range_id`
* `diameter`
* `date_planted`
* `longitude`
* `latitude`

```{r}
vars_of_interest <- c("tree_id", "std_street", "common_name", "height_range_id", "diameter", "date_planted", "longitude", "latitude")

vancouver_trees %>% select(vars_of_interest)
```

From we can assess what the value in each variable is representing.

| Variable | Value |
| --- | --- |
| `tree_id` | The ID of a given tree |
| `std_street` | The street where a given tree is located |
| `common_name` | The common name of the species/cultivar of a tree |
| `height_range_id` | Categorized height of a given tree |
| `diameter` | Diameter across the trunk of a given tree |
| `date_planted` | The date when a given tree was planted |
| `longitude` | The longitude of a given tree |
| `latitude` | The latitude of a given tree |

## 2.2 Untidying my data

I can untidy my data by using `pivot_longer` to create a new variable called `location_metric` that will either be `longitude` or `latitude` and the value will be the longitudinal or latitudinal coordinate of the tree. This will untidy the data since there will be two rows per tree

```{r}
(untidy_vancouver_trees <- vancouver_trees %>% 
   select(vars_of_interest) %>% 
   pivot_longer(cols = c(longitude, latitude),
                names_to = "location_metric", 
                values_to = "coordinate")) %>% 
  select(location_metric, coordinate, everything())
```

Tidy this back up with `pivot_wider` to get one row per tree again.

```{r}
untidy_vancouver_trees %>% 
  pivot_wider(names_from = location_metric, values_from = coordinate) %>% 
  select(longitude, latitude, everything())
```

## 2.3 Picking 2 research questions

I would like to continue exploring research questions 1 and 4. 

### Research Question 1

*How has the frequency of trees planted by neighbourhood changed over time? Which neighbourhoods saw the greatest rate of trees planted per decade? Are there any neighbourhoods that have only recently seen a high rate of tree planting (last 5 years?) and conversely, are there any neighbourhoods that have drastically slowed down in their rate of tree planting?*

This question mainly interests me because of the potential it has to serve as a signal of development in neighbourhoods.There are many different angles from which I can approach this question. 

I want to explore the refined version of this question, looking at the relative rate of trees planted per decade in each neighbourhood. 

To answer this question, I need to first filter out observations that do not have `date_planted` information.

```{r}
# keep modified dataset to answer question 1 as q1_data
(q1_data <- vancouver_trees %>% 
  filter(!is.na(date_planted)))
```

I only really need neighbourhood and date information in order to answer this question, so I can just select the relevant variables.

```{r}
(q1_data <- q1_data %>% 
  select(neighbourhood_name, date_planted))
```

Since I'm interested in looking at the annual rate of planting, I'll create a `year_planted` variable.

```{r}
(q1_data <- q1_data %>% 
  mutate(year_planted = lubridate::year(date_planted)))
```

Rather than looking at the data from a tree-by-tree basis, I'd rather explore by neighbourhood. 

```{r}
(q1_data <- q1_data %>% 
  group_by(neighbourhood_name, year_planted) %>% 
  summarise(trees_planted = n()))
```

I would like 1 row per neighbourhood, with each annual number of trees planted as a variable.

```{r}
(q1_data <- q1_data %>% 
  pivot_wider(names_from = year_planted, values_from = trees_planted))
```

There are a few NA values, these are years where no trees were recorded, replace the NA values with 0

```{r}
(q1_data <- q1_data %>% 
  replace(is.na(.), 0))
```

### Research Question 4 (Now referred to as question 2)

*Is there a relationship between the height and width of trees? Are these larger trees found in geographically similar areas or are they scattered throughout the city?*

I'd like to continue exploring this questions because I have not yet answered the second part of it. I'd also like to add a part to this question that explores the relationship between the age of a tree (based on `date_planted`) and the size of the tree.

I would like to only keep the relevant variables to the question, which are `height_range_id`, `diameter`, `latitude`, `longitude`, `date_planted`. It may be useful to have `on_street_block`, `on_street`, `root_barrier`, `common_name`, and `neighbourhood_name` as well for the exploration. 

```{r}
(q2_data <- vancouver_trees %>% 
  select(height_range_id, diameter, latitude, longitude, date_planted, on_street_block, on_street, root_barrier, common_name, neighbourhood_name))
```

Height range ID is currently a double type, since it is an ID value, it would make more sense to cast it to a factor. 

```{r}
(q2_data <- q2_data %>% 
   mutate(height_range_id = factor(height_range_id, 0:10)))
```

I'd also like to sort my data in order of increasing `height_range_id`.

```{r}
(q2_data <- q2_data %>% 
   arrange(height_range_id))
```

There are some trees that do not have location information or date planted information, filter those out.

```{r}
(q2_data <- q2_data %>% 
   filter(!is.na(longitude) | !is.na(latitude) | !is.na(date_planted)))
```

I may create some categorical variables based on diameter and date_planted, but will leave it for now. 




